A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-07-24 manuscript.qmd
add limitation of case study design to abstract and introduction

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
date: 2024-07-01
date-format: iso
author:
  - name: Florian Pargent
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - ref: lmu
  - name: Anne-Kathrin Kleine
    orcid: 0000-0003-1919-2834
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    orcid: 0000-0002-6600-9580
    roles:
      - conceptualization
      - funding acquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    orcid: 0000-0002-1633-4772
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: UCL Global Business School for Health
author-note:
  disclosures:
    data-sharing: This is version 3 of our preprint published at <https://doi.org/10.31234/osf.io/rpjem>. All materials (reproducible manuscript, including R code and all simulation results) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/dhwf4/>. A Quarto Manuscripts website is hosted at <https://Timo-Ko.github.io/glmm_simulation_tutorial/>.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525).
    gratitude: ~
    authorship-agreements: Florian Pargent and Timo Koch contributed equally to this manuscript.
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. The recent replication crisis underlines the need for more rigorous statistical methodology and adequately powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. However, available methods and software for a priori sample size planning for GLMMs are often limited to specific designs. Tailored data simulation approaches offer a more flexible alternative. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori sample size planning with GLMMs. We not only focus on power analysis but also explain how to use the precision of parameter estimates to determine appropriate sample sizes. We conclude with an outlook on the increasing importance of simulation-based sample size planning.
keywords: [tutorial, sample size planning, generalized linear mixed model, power analysis, precision, data simulation]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

When planning experimental research, it is essential to determine an appropriate sample size and use appropriate statistical models to analyze the data to ensure that the results are robust and informative [@lakensSampleSizeJustification2022]. The recent replication crisis in Psychology and other disciplines has illustrated many challenges surrounding the reproducibility and reliability of study findings [@yarkoniGeneralizabilityCrisis2022]. As a result, there is a growing need for more rigorous statistical methodology and the adoption of adequately powered experimental designs. Multiple easy-to-use software solutions exist for simple statistical models and experimental designs [@R-pwr; @lakensSimulationBasedPowerAnalysis2021]. A priori sample size planning for more complex research designs such as flexible generalized linear mixed models (GLMM) is not covered by standard software solutions. Researchers willing to use this framework will need to use data simulation. In the present work, we provide a tutorial on how to determine adequate sample sizes by performing tailored simulation-based sample size planning for GLMMs. After introducing some theoretical background on sample size planning, we review existing software solutions in R and discuss under which circumstances tailored data simulations are necessary. We proceed by describing the relevant steps and decisions involved in tailored data simulation, illustrated in a case study.

>>>>

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
date: 2024-07-01
date-format: iso
author:
  - name: Florian Pargent
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - ref: lmu
  - name: Anne-Kathrin Kleine
    orcid: 0000-0003-1919-2834
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    orcid: 0000-0002-6600-9580
    roles:
      - conceptualization
      - funding acquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    orcid: 0000-0002-1633-4772
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: UCL Global Business School for Health
author-note:
  disclosures:
    data-sharing: This is version 3 of our preprint published at <https://doi.org/10.31234/osf.io/rpjem>. All materials (reproducible manuscript, including R code and all simulation results) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/dhwf4/>. A Quarto Manuscripts website is hosted at <https://Timo-Ko.github.io/glmm_simulation_tutorial/>.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525).
    gratitude: ~
    authorship-agreements: Florian Pargent and Timo Koch contributed equally to this manuscript.
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. The recent replication crisis underlines the need for more rigorous statistical methodology and adequately powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. However, available methods and software for a priori sample size planning for GLMMs are often limited to specific designs. Tailored data simulation approaches offer a more flexible alternative. Based on a practical case study where we focus on a binomial GLMM with two random intercepts and discrete predictor variables, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori sample size planning with GLMMs. We not only focus on power analysis but also explain how to use the precision of parameter estimates to determine appropriate sample sizes. We conclude with an outlook on the increasing importance of simulation-based sample size planning.
keywords: [tutorial, sample size planning, generalized linear mixed model, power analysis, precision, data simulation]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

When planning experimental research, it is essential to determine an appropriate sample size and use appropriate statistical models to analyze the data to ensure that the results are robust and informative [@lakensSampleSizeJustification2022]. The recent replication crisis in Psychology and other disciplines has illustrated many challenges surrounding the reproducibility and reliability of study findings [@yarkoniGeneralizabilityCrisis2022]. As a result, there is a growing need for more rigorous statistical methodology and the adoption of adequately powered experimental designs. Multiple easy-to-use software solutions exist for simple statistical models and experimental designs [@R-pwr; @lakensSimulationBasedPowerAnalysis2021]. A priori sample size planning for more complex research designs such as flexible generalized linear mixed models (GLMM) is not covered by standard software solutions. Researchers willing to use this framework will need to use data simulation. In the present work, we provide a tutorial on how to determine adequate sample sizes by performing tailored simulation-based sample size planning for GLMMs. After introducing some theoretical background on sample size planning, we review existing software solutions in R and discuss under which circumstances tailored data simulations are necessary. We proceed by describing the relevant steps and decisions involved in tailored data simulation, illustrated in a case study where we focus on a binomial GLMM with two random intercepts and discrete predictor variables.
